---
title: "R Fundamentals 9 - Making your own functions"
author:
   - name: Andrew Moles
     affiliation: Learning Developer, Digital Skills Lab
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    theme: readable
    highlight: pygments
    keep_md: yes
    code_download: true
    toc: true
    toc_float: true
---

# Objective of workshop

To start making your own functions to use for your data cleaning and analysis in R. 

# What this workshop will cover

In this workshop, the aim is to cover how to start making your own functions. We will be covering:

-   Introduce the basics of making functions
-   How to use functions you've made
-   Some general use cases to get you started

------------------------------------------------------------------------



Ideas:

- For example or exercise, breathing function. Maybe example is triangle, exercise is adjust to square breathing
- How to source functions you've made from a script
- How to comment your functions
- When to enforce user input to functions, and when to add defaults
- At end, tidy evaluation to access columns when your functions use tidyverse functions


example of making a range function, which is the difference between min and max values. 
```{r}
library(palmerpenguins)

range2 <- function(x, rem.na = TRUE) {
  range <- max(x, na.rm = rem.na) - min(x, na.rm = rem.na)
  return(range)
}

range2(penguins$body_mass_g)
```


Function to tell you what *generation* you are in.
```{r}
generation <- function(birth_year){
  if(birth_year >= 1997){
    gen <- "Gen Z"
  }
  else if(birth_year <= 1996 & birth_year >= 1981){
    gen <- "Millennials"
  }
  else if(birth_year <= 1980 & birth_year >= 1965){
    gen <- "Gen X"
  }
  else if(birth_year <= 1964 & birth_year >= 1955){
    gen <- "Boomers II"
  }
  else if(birth_year <= 1954 & birth_year >= 1946){
    gen <- "Boomers I"
  }
  else if(birth_year <= 1945 & birth_year >= 1928){
    gen <- "Post War"
  }
  else if(birth_year <= 1927 & birth_year >= 1922){
    gen <- "WW II"
  }
  else{
    gen <- "Not sure"
  }
  return(gen)
}

generation(1989)
```

