---
title: "R Workshop 6 - Tidyverse introduction with Pipes and dplyr"
author:
   - name: Andrew Moles
     affiliation: Learning Developer, Digital Skills Lab
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    theme: readable
    highlight: pygments
    keep_md: yes
    code_download: true
    toc: TRUE
    toc_float: TRUE
---

Packages to be used: magrittr, dplyr, tidyr, and readr (just for data loading)

Nice reference point - https://jhudatascience.org/tidyversecourse/index.html

# What is the tidyverse?



The tidyverse is a collection of R packages that are designed for data science. These packages share design, syntax, and philosophy. These packages cover the import of data (`readr` and `haven`), manipulation and transformation of data (`dplyr`, `tidyr`, `stringr`, `purrr`, `forcats`, and `lubridate`), visualisation (`ggplot` and it's extensions), and analysis (`tidymodels`). 

Essentially, the tidyverse makes data science in R less painless, improving your experience of R and data science, especially in the data cleaning and wrangling stages. 

# What is tidy data? 

The tidyverse has a focus on working with tidy data, or making data tidy, ready for visualisation and analysis. So what does tidy data mean?

When your data is tidy, *each column is a variable*, *each row is an observation*, and *each cell is a single observation*, as per our example below:
```{r}
# tidy data example
tidy_df <- data.frame(
  id = 1:6,
  name = c("floof", "max", "cat", "donut", "merlin", "panda"),
  colour = c("grey", "black", "orange", "grey", "black", "calico")
)

tidy_df
```

Messy data is inconsistent and unique, making it harder to work with, and harder for others to work with. See this example of a messy dataset that would be hard to work with. 

```{r}
# example messy_df
messy_df <- data.frame(
  id = c(1,1,2,2,3,3,4,4,5,5,6,6),
  animal = c("floof", "grey",
             "max", "black",
             "cat", "orange",
             "donut", "grey",
             "merlin", "black",
             "panda", "calico")
)

messy_df
```

See this excellent article, which has lots of nice images, for a summary :-https://www.openscapes.org/blog/2020/10/12/tidy-data/

# Package install task

In this workshop we will be using four packages: magrittr, dplyr, tidyr, and readr. 

Using the code chunk below, install all four of these packages. *note that dplyr is large and might take a minute or so to install*
```{r eval=FALSE}
# your code here
install.packages("")
install.packages("")
install.packages("")
install.packages("")
```

# Intro to pipes

The pipe operator in R comes from the `magrittr` package, using syntax of %>%. 

The pipe operator is for chaining a sequence of operations together. This has two main advantages: it makes your code more readable, and it saves some typing. 

Run the example below to see the difference between using pipes and not using pipes. We are going to calculate a mean of a vector of numbers, then round the result. First we look at what we do when we don't use pipes, which you might be familiar with. 

```{r}
# load in magrittr
library(magrittr)

# Make some data: 20 randomly selected data points, from 1 to 10
x <- sample(1:10, 20, replace = TRUE)
y <- sample(1:10, 20, replace = TRUE)

# without pipe
y_mean <- mean(y)
y_mean <- round(y_mean, digits = 2)
y_mean <- paste("Mean value of y is", y_mean)
y_mean
```

Now lets have a look at how to do this same operations with a pipe. There are two ways, but both work the same. You can either assign the result at the beginning of your pipe operation, or at the end, as shown in the examples. 

You will notice in the paste function we have used a `.` after the text. This is called a *placeholder* whereby instead of using x (like we did above without the pipe) we add a . to tell R that is where we want our data to go. 

```{r}
# magrittr pipe
x_mean <- x %>% # assign result at the start
  mean() %>% 
  round(digits = 2) %>%
  paste("Mean value of x is", .) # we use the . as a place holder for a variable (e.g. instead of x)

x_mean

# a pipe can also be written like this 
x %>%
  mean() %>%
  round(digits = 2) %>%
  paste("Mean value of x is", .) -> x_mean # assign result at the end

x_mean
```

It is also worth mentioning that as of version 4.1 of R, base R comes with a native pipe operator. This has just been introduced, and may get more use in examples you'll see online in the future. The syntax uses |> as the pipe, and the structure is the same as a magrittr pipe. 

*note that the native pipe currently doesn’t have a placeholder, so we won't use paste in this example*
```{r}
# native R pipe
z <- sample(1:10, 20, replace = TRUE)

z_mean <- z |> # assign result at the start
  mean() |>
  round(digits = 2)

z_mean

z |>
  mean() |>
  round(digits = 2) -> z_mean # assign result at the end

z_mean
```

We will be using the magrittr pipe for the rest of this workshop, as it's currently the pipe operator you will come across most in the r world. 

## Exercise - using pipes

Using the vector of temperature provided and using magrittr pipes:

1) Pipe median and paste functions together to print the message: "median temp is 15"   
2) Pipe max and paste functions together to print the message: "max temp is 20" 

*hint: don't forget to use the placeholder with paste*
```{r}
library(magrittr)

temperature <- c(10, 16, 12, 15, 14, 15, 20)

# your code here
temperature %>%
  median() %>%
  paste("median temp is", .)
  
temperature %>%
  max() %>%
  paste("max temp is", .)

```

# Introduction to dplyr

Dplyr is a package that is built for data manipulation, using functions that describe what they do. For example, the `select()` function selects columns you want or don't want from a data frame. 

The dplyr package has a lot of functions built into the package, each has it's own very helpful documentation page with examples - https://dplyr.tidyverse.org/reference/index.html

Dplyr functions work with and without pipes and you'll see both when searching online. We will try use examples of both ways. 

We will focus on the key dplyr functions for now: `select()`, `filter()`, and `mutate()`. We will use the messi career data as for the examples. 

```{r}
# create the messi career data
messi_career <- data.frame(Appearances = c(9,25,36,40,51,53,55,60,50,46,57,49,52,54,50,44),
                           Goals = c(1,8,17,16,38,47,53,73,60,41,58,41,54,45,51,31),
                           Season = c(2004,2005,2006,2007,2008,2009,2010,2011,2012,
            2013,2014,2015,2016,2017,2018,2019),
                           Club = rep("FC Barcelona", length(16)),
                          Age = seq(17, 32),
                          champLeagueGoal = c(0,1,1,6,9,8,12,14,8,8,10,6,11,6,12,3))
# view the data
head(messi_career)
```

## Select function

The select function subsets columns from a data frame using their name. There are several different ways of using select. Run each of the code chunks below and review the outputs. 

```{r message=FALSE, warning=FALSE}
# load dplyr
library(dplyr)

# select single column
messi_career %>% select(Goals)
```

```{r}
# select multiple columns
messi_career %>% select(Appearances, Goals, Age)
```

```{r}
# select slice (or range) of columns
messi_career %>% select(Season:Age)
```

```{r}
# negate selection of column
messi_career %>% select(!(Season:Age))
```

```{r}
# select slice and other columns
messi_career %>% select(Appearances:Season, champLeagueGoal)
```

```{r}
# select by literal string
messi_career %>% select(contains("Goal"))
```

```{r}
# negate selection with slice and extra column (note c() function used)
messi_career %>% select(!c(Season:Age, champLeagueGoal))
```

As you can see, `select()` makes it easy to extract columns from your data, and becomes more useful the larger your dataset becomes. 

In the examples above we did not assign the result. See the examples below on how to do this. 
```{r}
# assign result to subset
messi_sub <- messi_career %>%
  select(contains("Goal"))

messi_sub

# The no pipe method
messi_sub <- select(messi_career, contains("Goal"))
```

# Select exercise 

For your exercises, you will be using imdb movie data! I've loaded it here in the code for you. 

The data has 22 columns, some of which we won't need. We can use select to subset our data to keep only what we want. 

Using select with pipes, subset the `imdb_movie` data so you have the following columns:  
- imdb_id through to writer
- actors
- avg_vote to votes
- reviews_from_users to reviews_from_critics

Assign the result to `imdb_sub`. 

Use glimpse to review the subsetted data: *data %>% glimpse()*

*hint: you should be able to fit this into one select call*
```{r message=FALSE, warning=FALSE}
# load libraries
library(readr)
library(dplyr)

# load data
movies_imdb <- read_csv("https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/Workshop6/data/IMDb%20movies.csv")

# use glimpse to review data (tidyverse version of str())
movies_imdb %>% glimpse()

# your code here
imdb_sub <- movies_imdb %>%
  select(imdb_title_id:writer, actors, avg_vote:votes, reviews_from_users:reviews_from_critics)

imdb_sub %>% glimpse()
```

There is a more efficient way of doing this using select. From looking at the examples provided, can you think of a better way of taking out the columns we removed?

# Filter function

The filter function allows you to subset rows based on conditions, using conditional operators (==, <=, != etc.). It is similar to the base r `subset()` function which we have used in previous R workshops. The table below is a reminder of the conditional operators you can use. 

| Operator | Meaning                  |
|----------|--------------------------|
| >        | Greater than             |
| >=       | Greater than or equal to |
| <        | Less than                |
| <=       | Less than or equal to    |
| ==       | Equal to                 |
| !=       | Not equal to             |
| !X       | NOT X                    |
| X | Y    | X OR Y                   |
| X & Y    | X AND Y                  |
| X %in% Y | is X in Y                |


Just like when using `select`, you provide the column name you want to apply conditional logic to. If you are piping, you don't need to provide your data as an argument in the function. 
```{r}
# filter based on one criteria
messi_career %>% filter(Goals > 50)

# filter then pipe to select
messi_career %>% filter(Appearances >= 55) %>%
  select(Season, Age)

# filter on more than one condition
messi_career %>% filter(Goals > 50 & champLeagueGoal <= 10)

# filter on average
messi_career %>% filter(Goals > mean(Goals, na.rm = TRUE))
```

To assign the result to a new data frame (subset) we use the assignment operator as the beginning or the end of our code; here we have just shown the beginning, in the pipes section we show both versions.
```{r}
# assign result to messi_sub
messi_sub <- messi_career %>%
  filter(Appearances <= 40) %>%
  select(Goals, Age)

# view result
messi_sub
```

## Filter exercise

We are going to filter our subsetted (`imdb_sub`) data to find the best rated films from the USA in the year 1989, and create a subset called USA_1989_high.

1) Pipe from imdb_sub to filter, filtering country being equal to USA
2) Pipe from your country filter to another filter, filtering for year being equal to 1989
3) Pipe from your year filter to your final filter. Filter for avg_vote to be greater than or equal to 7.5 and reviews_from_critics to be greater than 10
4) Make sure to assign your result to USA_1989_high
5) Print the result to see the highest rated films, made in the USA, in 1989. 

```{r}
# your code here

USA_1989_high <- imdb_sub %>%
  filter(country == "USA") %>%
  filter(year == 1989) %>%
  filter(avg_vote >= 7.5 & reviews_from_critics > 10)

USA_1989_high
```

You might have noticed that the country column has some strings that are split by a comma, e.g. Germany, Denmark. The == operator will not be able to pick these up. Instead we would use the base R `grepl()` function or `str_detect()` from the `stringr` package. This won't be covered in this workshop, but will be in future workshops. If you are interested, have a look at the stringr package - https://stringr.tidyverse.org/index.html. 

# Mutate function

The mutate function is for making, modifying, or deleting columns in your dataset. Similar to what we have done in previous sessions, mutate allows you to make a new column from a calculation you have made. 

The main difference between using matate and making new columns in base R, is that mutate is smarter. You can create a new column based on a new column you have just made in mutate, which you can't do in base R. Lets look at some examples with our messi data. 

In our previous workshops, we calculated Messi's goals per game (goals/appearances). We can do this with mutate. Notice the syntax, we give the name we want to call our new column first, then =, then what we want to do (e.g. a calculation).

```{r}
messi_career %>%
  mutate(goal_ratio = Goals/Appearances)
```
The new column, goal_ratio in this case, will automatically be added to the end of your data.frame. This is the same behaviour you will see when using base R. 

What makes `mutate()` powerful, is the ability to do multiple calculations in one statement, as well as using newly made columns. See the below example which will help understand this. We will use goal ratio to find out the difference between goal ratio and the average goal ratio for each row (or season). 

```{r}
# calculate goal ratio and diff from mean
messi_career %>%
  mutate(
    goal_ratio = round(Goals/Appearances, digits = 2),
    diff_avg_goal_ratio = goal_ratio - (mean(Goals) / mean(Appearances))) -> messi_career

# print result
messi_career
```

We can then pipe this result to `filter()`, which allows us to see which seasons Messi has a goal ratio above his average goal ratio. 
```{r}
messi_career %>%
  mutate(
    goal_ratio = round(Goals/Appearances, digits = 2),
    diff_avg_goal_ratio = goal_ratio - (mean(Goals) / mean(Appearances))) %>%
  filter(diff_avg_goal_ratio > 0)
```

## Mutate exercise

Lets pretend we are interested in the difference between the number of user reviews and critic reviews for each film in our USA_1989_high dataset we just made in the last task. We can use mutate to explore this difference a bit further. 

1)  Pipe your USA_1989_high data to a `mutate()` function. Make a new column called `user_critic_ratio`, and divide reviews_from_users by reviews_from_critics. Wrap the result in a `round()` function, rounding by two digits
2)  Now pipe to a `select()` function, selecting out the title, avg_vote and user_critic_ratio columns
3)  Now pipe to a `filter()` function, filtering for user_critic_ratio to be greater than 4. 

You should get a data frame returned that has the films The Abyss, Dead Poets Society, Do the Right Thing, and Glory.

We are going to try and re-use the user_critic_ratio variable we made. Adding to the code you just made:
4)  In your mutate function, add another column called `ratio_string`. Using the `paste()` function, use the title column and your user_critic_ratio column to make a string that looks like "The Abyss has a user to critic ratio of 4.08"
5)  Make sure to add your `ratio_string` column to your select function

You should now see the ratio_string column you made next to your user_critic_ratio column in the output. 

```{r}
# your code here

# user critic review ratio
USA_1989_high %>%
  mutate(user_critic_ratio = round(reviews_from_users / reviews_from_critics, digits = 2)) %>%
  select(title, avg_vote, user_critic_ratio) %>%
  filter(user_critic_ratio > 4)

# put ratio intro a string
USA_1989_high %>%
  mutate(user_critic_ratio = round(reviews_from_users / reviews_from_critics, digits = 2),
         ratio_string = paste(title, "has a user to critic ratio of", user_critic_ratio)) %>%
  select(title, avg_vote, user_critic_ratio, ratio_string) %>%
  filter(user_critic_ratio > 4)

```
We can see we get more user reviews than critic reviews, which makes sense; for example, the The Abyss has 4 user reviews for each critic review. 

# Mutate with the across function

We can take the mutate function further by using the across function. This allows us to perform operations (do something) across multiple columns. This is very useful for doing type conversions in an efficient way. 

The across function works in a similar way to the `select` function, but if you want to pick out a few columns you have to use the `c()` function. See the examples below, where we have selected two columns, or used a slice to select out a few columns that are next to each other.
```{r}
# perform round (to 1 decimal place) across selected columns
messi_career %>%
  mutate(across(c(goal_ratio, diff_avg_goal_ratio), round, digits = 1))

# square root across columns selected with slice
messi_career %>%
  mutate(across(1:3, sqrt))

# square root across columns selected with slice (using col names)
messi_career %>%
  mutate(across(Appearances:Season, sqrt))
```

We can also combine the across function with the `where()` or `all_of()` function to perform conditional mutations. 

The `where()` function does conditional matching between the statement you've used and what is in your dataset. In the example we are asking `where()` to look for columns that are the character data type. Then we can perform an operation, such as convert those columns to factors. 

The `all_of()` function looks for matches between the strings you have provided and the column names in your dataset. In our example, we put the Season and Club columns into a vector, then call that vector and convert them to a factor.
```{r}
# perform conditional operation with where
messi_career %>%
  mutate(across(where(is.character), as.factor)) %>%
  glimpse()

# change selected variables with all_of
to_factor <- c("Season", "Club")

messi_career %>%
  mutate(across(all_of(to_factor), as.factor)) %>%
  glimpse()
```

## Across function exercise

Lets go back to our imdb_sub data. We want to extract films from 1990 through to 1995, that are from the USA, and have an avg_vote greater than or equal to 7.5. 

Let's also say we want all our variables that are currently characters to be factors, and want the year column to also be a factor. 

1) Using the imdb_sub data, filter for years between and including 1990 and 1995
2) Now also filter for the country to be the USA, with an avg_vote greater then or equal to 7.5
3) Using mutate, across and where, convert any column that has a character data type to a factor
4) Using mutate, convert year to a factor
5) Save the result in a data frame called `USA_early90_high`
6) Using your new USA_early90_high subset, filter for avg_vote greater than or equal to 8.5, then select the title, avg_vote, and year columns. View the result to see the top rated films and what year they were in. 

```{r}
# your code here

# first way of doing this
imdb_sub %>% 
  filter(year >= 1990 & year <= 1995) %>%
  filter(country == "USA" & avg_vote >= 7.5) %>%
  mutate(across(where(is.character), as.factor),
         year = as.factor(year)) -> USA_early90_high

# second way of doing this
to_factor <- c("year")

imdb_sub %>% 
  filter(year >= 1990 & year <= 1995) %>%
  filter(country == "USA" & avg_vote >= 7.5) %>%
  mutate(across(where(is.character), as.factor),
         across(any_of(to_factor), as.factor)) -> USA_early90_high

# highest rated with title, avg_vote, and year
USA_early90_high %>%
  filter(avg_vote >= 8.5) %>%
  select(title, avg_vote, year)

```

