---
title: "R Workshop 5 - Conditionals and Logic"
author:
   - name: Andrew Moles
     affiliation: Learning Developer, Digital Skills Lab
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    theme: readable
    highlight: pygments
    keep_md: yes
---

# What this workshop will cover
*  Boolean operators 
*  Conditional operators
*  Conditional logic
*  If else statements

## Why this style?

*  Online training is tiring so keeping the sessions to one hour
*  No or limited demonstrations provided in order to provide more real world experience - you have a problem and you look up how to solve it, adapting example code
*  Trainer support to guide through process of learning

## We will be working in pairs:

*  Option to work together on worksheet or to work individually
*  If possible have your camera on and introduce yourself to each other

## What to do when getting stuck:

1)  Ask your team members
2)  Search online:
  *  The answer box on the top of Google's results page 
  *  stackoverflow.com (for task-specific solutions)
  *  https://www.r-bloggers.com/ (topic based tutorials)
3)  Don't struggle too long looking online, ask the trainer if you can't find a solution!

***

# Boolean operators

In these workshops we have covered numeric, string, and factorised data types. The one data type we have not yet covered in this workshop is the logical data type. The logical data type is boolean; boolean is a binary variable that can have one of two possible values, 0 (false) or 1 (true).

To use boolean in R we capitalise true and false. Run the code below to test this out. 
```{r}
trueFalse <- c(TRUE, FALSE, TRUE, FALSE)
class(trueFalse)
```

We can also convert binary into boolean. R will make a 1 true and a 0 false. Run the code below to test this out. 
```{r}
binary <- c(1,0,0,0,1,1,1,0)
as.logical(binary)
```

Why is this useful? Using boolean allows us to set *conditions* that can either be true or false. Meaning we can test variables against each other on the basis of a condition being met with the outcome being determined by the test. 

## Boolean task

1) Make data frame called pizza. For the first column use the topping vector provided. For the second column call it Good_topping, and use 1's and 0's to indicate a good or bad topping (1 = good, 0 = bad)
2) Convert your good topping column to logical using `as.logical`
3) Run `str()` and `head()` to view your data. Good_topping should now be logical (true and false). 

```{r}
Topping <- c('Mozzarella', 'Pepperoni','Pinapple', 'Pepper', 'Carrot')

# your code here
pizza <- data.frame(
  Topping,
  Good_topping = c(1,1,0,1,0)
)

pizza$Good_topping <- as.logical(pizza$Good_topping)

str(pizza)
```

# Conditional Operators

In order to test conditions we need conditional operators. Below is a table of conditional operators. During this session we will have examples and tasks using them in various contexts. 

| Operator | Meaning                  |
|----------|--------------------------|
| >        | Greater than             |
| >=       | Greater than or equal to |
| <        | Less than                |
| <=       | Less than or equal to    |
| ==       | Equal to                 |
| !=       | Not equal to             |
| !X       | NOT X                    |
| X | Y    | X OR Y                   |
| X & Y    | X AND Y                  |
| X %in% Y | is X in Y                |

The greater than (>), or greater than or equal to (>=) operators test if variable x is greater than y; less than (<), or less than or equal to (<=) test if x is less than, or equal to, y. The output is always boolean. 

Run the code below to test this out. 
```{r}
18 > 18
18 >= 18
18 < 17
```
Why is 18 > 18 false? Any value below 18 would be true, but 18 is not greater than 18. To include 18 into the test we have to use the greater than or equal to operator. 

You're most likely to use these tests on data frames or vectors. Run the code below to test out using equals to and not equal to on two columns in a data frame. 
```{r}
# data frame to test on
df <- data.frame(
  x = c(rep(20,5),rep(40,2),rep(60,3)),
  y = c(rep(20,3),rep(40,5),rep(60,2))
)
# view data
df
# where is x equal to y
df$x == df$y
# where is x not equal to y
df$x != df$y
# we can also count the true cases
sum(df$x == df$y)
```

The *equal to* test tells us where x and y have the same data, *not equal to* does the opposite. We can also use conditional operators on strings. Run the code below to test this out. 
```{r}
"pizza" == "bad"
"pizza" == "pizza"
```

## Conditional operator task 

Using conditional operators work out the following:

1) Is 70 less than or equal to 11?
2) Using the following vector, `z <- sample(1:100, 20)`, count how many elements in *z* are less than (<) 50, and how many are greater then or equal to (>=) 50. 
3) Using the following vector, `pets <- c(rep('cat',5), rep('fish',11), rep('dog',6))`, count how many fish there are.
4) Using the same pets vector, how many pets are not fish?

```{r}
# your code here
# 1 
70 <= 11
# 2
z <- sample(1:100, 20)
sum(z < 50)
sum(z >= 50)
# 3
pets <- c(rep('cat',5), rep('fish',13), rep('dog',6))
sum(pets == 'fish')
# 4
sum(pets != 'fish')
```

# Indexing using conditional operators (subsetting)

Indexing using conditional operators is the same as we learned in previous sessions. Instead of calling the index of vector or data frame, we call a condition to test and find data. 

We might want to find specific data, in the example below we are subsetting data from the pets vector that is equal to cat. Run the code to test it out. 
```{r}
# pet data
pets <- c(rep('cat',5), rep('fish',13), rep('dog',6))
# pull out cats
cats <- pets[pets == 'cat']
# view subsetted data
cats
```

When doing the same with data frames you will need to include a comma. The condition goes in the row index (left hand of comma); Remember `df[row index, col index]`. Run the code below to test it out on data frames. 
```{r}
# data frame with numbers randomly pulled from 1 to 100 for x & y
df <- data.frame(
  x = sample(1:100, 10),
  y = sample(1:100, 10)
)
# showing data where x is less than 20
df[df$x < 20, ]
# subsetting data into df2 that are >= 35
df2 <- df[df$x >= 35, ]
df2
```

In base R there is a specialised function for subsetting data frames called `subset()`. It is pretty handy as it will save you time typing because you only need to give the name of the data frame once. 
```{r}
subset(df, y <= 40)
```

## Subsetting task

In this task we are going to load in data from a URL and do some subsetting with it. The dataset is from the Pokémon games, and includes only the original game Pokémon (Pikachu, Mewtwo etc.). We will use this data for the rest of the tasks. Each row in the data is a different Pokémon, with their various statistics and typing. 

1) Using `read_csv()` from the `readr` library load in the data from the following URL, calling the data pokemon: https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/Workshop4b/data/pokemonGen1.csv
2) Get information on your loaded data using the `str()`, `head()` and `View()` functions. 
3) Using `subset()` make a new data frame called highHP and subset pokemon with a HP greater than or equal to 80. *HP stands for hit points*
4) Run `summary()` on your highHP data frame to see the statistics of pokemon with high hit points (HP)
5) There are some very low stats for Attack, Defense, and Speed. Use `which.min()` to find out which pokemon have these stats. *hint: data[which.min(data$col),]*. 
6) The attack stat seems to have the highest mean (other than HP). Using `sum()` find out how many of your high HP pokemon have an Attack stat greater than or equal to 100.
7) Finally, find out who those pokemon are. *hint: use subsetting*

```{r}
# load in readr (install using install.packages('readr') if you don't have it
library(readr)

# your code here
# load in data
pokemon <- read_csv("https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/Workshop4b/data/pokemonGen1.csv")
# get information about dataset
head(pokemon)
str(pokemon)
# subset based on hit points
highHP <- subset(pokemon, HP >= 80)
summary(highHP)
# find low attack and defence
highHP[which.min(highHP$Attack),]
highHP[which.min(highHP$Defense),]
highHP[which.min(highHP$Speed),]
# high attack
sum(highHP$Attack >= 100)
subset(highHP, Attack >= 100)
```

# Using AND OR and %in%

So far we have only been using one condition. By using the AND/OR operators we can thread together multiple conditional tests to make our code more efficient. 

Both statements behave slightly differently. When using AND you must satisfy both conditional tests, when using OR you can satisfy either of your conditional tests. Therefore, AND is strict, while OR is more lenient. 

For example, we have the 7 times table from -70 to 140 and we want to find what numbers are below 20 but above 0. AND works really well for this as it is strict, however OR doesn't work so well and includes everything. OR is better used if we wanted to find numbers less than 0 OR over 70. 
```{r}
# 7 times table number sequence 
s <- seq(-70, 140, by = 7)
s
# AND 
s[s < 20 & s > 0]
# OR
s[s < 20 | s > 0]
# OR finding each end of seq
s[s < 0 | s > 70]
```

## AND OR exercise

Using the df3 data frame provided run the following tests. Be sure to run the rests using square brackets or the `subset()` function so you see the data as the output. 

1) Subset data where x is less than 50 but greater then 20
2) Subset data where y is greater than or equal to 70 and z is also greater than or equal to 70
3) Subset data where x, y, or z are all less than 15
4) Subset data where y is less than 60 and exp is equal to control
```{r}
# test data frame
df3 <- data.frame(
  id = paste("person_",letters[1:10]),
  x = sample(1:100, 10),
  y = sample(1:100, 10),
  z = sample(1:100, 10),
  exp = sample(c(rep("control",50),rep("test",50)),10)
)
# your code here
df3[df3$x < 50 & df3$x > 20,]
df3[df3$y >= 70 & df3$z >= 70,]
df3[df3$x < 15 | df3$y < 15 | df3$z < 15,]
subset(df3, y < 60 & exp == 'control')
```


# %in% operator
The %in% operator is a really handy way of finding data that matches a criteria you are looking for. 
```{r}
# test data frame
df3 <- data.frame(
  id = paste("person_",letters[1:10]),
  x = sample(1:100, 10),
  y = sample(1:100, 10),
  z = sample(1:100, 10),
  exp = sample(c(rep("control",50),rep("test",50)),10)
)

1:10 %in% df3$z
match <- paste("person_",letters[5:10])
subset(df3, id %in% match)

```
## AND OR %in% exercise



# The If statement

The if statement is fundamental to programming. It uses conditional operators and boolean to *control flow*. This means to run a test to determine which operation of output is given, so it controls the flow of your code. 

```{r}

```





