---
title: "R Workshop 7 - Data wrangling with dplyr continued"
author:
   - name: Andrew Moles
     affiliation: Learning Developer, Digital Skills Lab
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    theme: readable
    highlight: pygments
    keep_md: yes
    code_download: true
    toc: TRUE
    toc_float: TRUE
---

# What this workshop will cover

-  Data manipulation with mutate from dplyr
-  Renaming columns 
-  Cleaning up column names with janitor

## Why this style?

-   Online training is tiring so keeping the sessions to one hour
-   No or limited demonstrations provided in order to provide more real world experience - you have a problem and you look up how to solve it, adapting example code
-   Trainer support to guide through process of learning

## We will be working in pairs:

-   Option to work together on worksheet or to work individually
-   If possible have your camera on and introduce yourself to each other

## What to do when getting stuck:

1)  Ask your team members
2)  Search online:
-   The answer box on the top of Google's results page
-   stackoverflow.com (for task-specific solutions)
-   <https://www.r-bloggers.com/> (topic based tutorials)
3)  Don't struggle too long looking online, ask the trainer if you can't find a solution!

------------------------------------------------------------------------

# The mutate function

The mutate function is from the dplyr library, and is for making, modifying, or deleting columns in your dataset. Similar to what we have done in previous sessions, mutate allows you to make a new column from a calculation you have made.

The main difference between using mutate and making new columns in base R, is that mutate is smarter. You can create a new column based on a new column you have just made within mutate, which you can't do in base R. Lets look at some examples with our messi data we used in the last session.

In our previous workshops, we calculated Messi's goals per game (goals/appearances). We can do this with mutate. Notice the syntax, we give the name we want to call our new column first, then =, then what we want to do (e.g. a calculation);  `mutate(new_column = x/y)`.

*note: when loading dplyr you also load the magrittr library for piping*
```{r message=FALSE, warning=FALSE}
# load dplyr
library(dplyr)

# create the messi career data
messi_career <- data.frame(Appearances = c(9,25,36,40,51,53,55,60,50,46,57,49,52,54,50,44),
                           Goals = c(1,8,17,16,38,47,53,73,60,41,58,41,54,45,51,31),
                           Season = c(2004,2005,2006,2007,2008,2009,2010,2011,2012,
            2013,2014,2015,2016,2017,2018,2019),
                           Club = rep("FC Barcelona", 16),
                          Age = seq(17, 32),
                          champLeagueGoal = c(0,1,1,6,9,8,12,14,8,8,10,6,11,6,12,3))
# view the data
head(messi_career)

# calculate the goal to appearance ratio
messi_career %>%
  mutate(goal_ratio = Goals/Appearances)
```

The new column, goal_ratio in this case, will automatically be added to the end of your data frame. This is the same behaviour you will see when using base R. This behaviour can be altered if you want, but we won't have time to cover it here. 

What makes `mutate()` powerful, is the ability to do multiple calculations in one statement, as well as using newly made columns. See the below example which will help to understand this. We will use goal_ratio to find out the difference between goal_ratio and the average goal ratio for each row (or season).

```{r}
# calculate goal ratio and diff from mean
messi_career <- messi_career %>%
  mutate(
    goal_ratio = round(Goals/Appearances, digits = 2),
    diff_avg_goal_ratio = goal_ratio - (mean(Goals) / mean(Appearances)))

# print result
messi_career
```

We can then pipe this result to `filter()`, which allows us to see which seasons Messi has a goal ratio above his average goal ratio.

```{r}
messi_career %>%
  mutate(
    goal_ratio = round(Goals/Appearances, digits = 2),
    diff_avg_goal_ratio = goal_ratio - (mean(Goals) / mean(Appearances))) %>%
  filter(diff_avg_goal_ratio > 0)
```

## Mutate exercise 1

We will be using the imdb movies dataset again for this workshop. Use the code below to load in the data.  

```{r message=FALSE, warning=FALSE}
# load libraries
library(readr)
library(dplyr)

# load data
movies_imdb <- read_csv("https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/Workshop6/data/IMDb%20movies.csv")

# use glimpse to review data (tidyverse version of str())
movies_imdb %>% glimpse()
```

Lets pretend we are interested in the difference between the number of user reviews and critic reviews for each film in our movies_imdb dataset. We can use mutate to explore this difference a bit further.

1)  Pipe your movies_imdb data to a `mutate()` function. Make a new column called `user_critic_ratio`, and divide `reviews_from_users` by `reviews_from_critics`. Wrap the result in a `round()` function, rounding by two digits
2)  Now pipe to a `filter()` function, filtering country to be USA and year to be 1989
2)  Now pipe to a `select()` function, selecting the title, avg_vote and user_critic_ratio columns
3)  Now pipe to a `slice_max` function, extracting rows that had the top 10 avg_rating

You should get a data frame returned that has films including: The Abyss, Dead Poets Society, Do the Right Thing, and Glory.

```{r}
# your code here

# user critic review ratio
movies_imdb %>%
  mutate(user_critic_ratio = round(reviews_from_users / reviews_from_critics, digits = 2)) %>%
  filter(country == "USA" & year == 1989) %>%
  select(title, avg_vote, user_critic_ratio) %>%
  slice_max(avg_vote, n = 10)
```

We can see we get more user reviews than critic reviews, which makes sense; for example, the The Abyss has 4 user reviews for each critic review.

## Mutate exercise 2

In our second mutate exercise, you will need to de-bug the code to get it running! You may need to re-order some elements of the code as well as checking for other errors. 

We are filtering the movies_imdb data for films that are from the USA before the year 1990, have a duration less than 120 minutes, and an average vote greater than 8.5. We will also be using the user_critic_ratio column to make it into a string for easier reading. 

You should end up with a data frame with 6 rows, and 4 columns (title, year, avg_vote, and ratio_string). The final column, ratio_string, should have an output like "Psycho has a user to critic ratio of 5.44". 

```{r eval=FALSE}
# your code here
usa_pre90_high <- movies_imdb |>
  mutate(user_critic_ratio = round(reviews_from_users / reviews_from_critics, digits = 2),
         ratio_string = paste(title, "has a user to critic ratio of", userCriticRatio)) %>%
  filter(country == "USA" & year < 1990) 
  select(title, year, avg_vote, ratio_string) %>%
  filter(duration < 120 & avg_vote >= 8.5)
  
usa_pre90_high
```

```{r}
# answer for solutions
# Re using the user_critic_ratio variable

usa_pre90_high <- movies_imdb %>%
  mutate(user_critic_ratio = round(reviews_from_users / reviews_from_critics, digits = 2),
         ratio_string = paste(title, "has a user to critic ratio of", user_critic_ratio)) %>%
  filter(country == "USA" & year < 1990) %>%
  filter(duration < 120 & avg_vote >= 8.5) %>%
  select(title, year, avg_vote, ratio_string)

usa_pre90_high
```

# Mutate with the across function

We can take the mutate function further by using the `across()` function. This allows us to perform operations (do something) across multiple columns. This is very useful for doing type conversions in an efficient way.

The across function works in a similar way to the `select()` function, but if you want to pick out a few columns you have to use the `c()` function. See the examples below, where we have selected two columns, or used a slice to select out a few columns that are next to each other.

```{r}
# perform round (to 1 decimal place) across selected columns
messi_career %>%
  mutate(across(c(goal_ratio, diff_avg_goal_ratio), round, digits = 1))

# square root across columns selected with slice
messi_career %>%
  mutate(across(1:3, sqrt))

# square root across columns selected with slice (using col names)
messi_career %>%
  mutate(across(Appearances:Season, sqrt))
```

We can also combine the across function with the `where()` or `all_of()` functions to perform conditional mutations.

The `where()` function does conditional matching between the statement you've used and what is in your dataset. In the example we are asking `where()` to look for columns that are the character (string) data type. Then we can perform an operation, such as convert those columns to factors. In this case it is just the Club column that changes. 

```{r}
# perform conditional operation with where
messi_career %>%
  mutate(across(where(is.character), as.factor)) %>%
  glimpse()
```

The `all_of()` function looks for matches between the strings you have provided and the column names in your dataset. In our example, we put the Season and Club columns into a vector, then call that vector and convert those columns to a factor.

```{r}
# change selected variables with all_of
to_factor <- c("Season", "Club")

messi_career %>%
  mutate(across(all_of(to_factor), as.factor)) %>%
  glimpse()
```

## Across function exercise

Lets go back to our movies_imdb data. We want to extract films from 1990 through to 1995, that are from the USA, and have an avg_vote greater than or equal to 7.5. We also want all our variables that are currently characters to be factors, and want the year column to also be a factor.

1)  Using the movies_imdb data, filter for years between and including 1990 and 1995
2)  Now also filter for the country to be the USA, with an avg_vote greater then or equal to 7.5
3)  Using mutate, across and where, convert any column that has a character data type to a factor
4)  Using mutate, convert year to a factor
5)  Save the result in a data frame called `usa_early90_high`
6)  Using your new `usa_early90_high` subset, filter for avg_vote greater than or equal to 8.5, then select the title, avg_vote, and year columns. View the result to see the top rated films and what year they were in.

```{r}
# your code here

# first way of doing this
usa_early90_high <- movies_imdb %>% 
  filter(year >= 1990 & year <= 1995 &
           country == "USA" & avg_vote >= 7.5) %>%
  mutate(across(where(is.character), as.factor),
         year = as.factor(year))

# second way of doing this
to_factor <- c("year")

usa_early90_high <- movies_imdb %>% 
  filter(year >= 1990 & year <= 1995 &
           country == "USA" & avg_vote >= 7.5) %>%
  mutate(across(where(is.character), as.factor),
         across(any_of(to_factor), as.factor))

# highest rated with title, avg_vote, and year
usa_early90_high %>%
  filter(avg_vote >= 8.5) %>%
  select(title, avg_vote, year)

```

# Ranking and cumulativate calculations using mutate

It can sometimes be helpful to rank your dataset, using mutate and the `min_rank()` or `percent_rank` functions allow you to add a new column with a rank based on a important column. Higher rank or percent rank means a better ranking. 

In this example, we want to make a goal ranking column and a percent raking column. We can then use filter to select rankings we are interested in. 
```{r}
messi_career <- messi_career %>%
  mutate(goal_rank = min_rank(Goals),
         goal_perc_rank = percent_rank(Goals))

# select rankings over 10
messi_career %>%
  filter(goal_rank > 10)
```

Another useful calculation you can do is to do cumulativate calculations, such as cumulativate sum or mean of a useful variable. For example, in our messi_career data it might be interesting to workout  his cumulativate goals, and average cumulativate appearances. We use the `cumsum()` and `cummean()` functions for these calculations. 

*note: cumulativate calculations are work very well with longitudinal data, like we have for Lionel Messi's career goals and appearances*

```{r}
messi_career %>%
  mutate(cumul_goals = cumsum(Goals),
         mean_cumul_app = cummean(Appearances)) %>%
  select(Goals, cumul_goals, Appearances, mean_cumul_app)
```

## Ranking and cumulativate calculations exercise

Using your usa_early90_high data we just made in the last exercise:

1)  Use mutate to make a new column called `duration_rank`, using the `min_rank()` function on the duration column
2)  In the same mutate statement, make a new column called `perc_duration_rank`, using the `percent_rank()` function on the duration column
3)  In the same mutate statement, make a new column called `avg_cumul_duration`, using the `cummean()` function on duration. 
4)  Pipe to a filter function, and filter for perc_duration_rank between 0.5 and 0.6
5)  Use select to extract the following columns: title, year, duration, avg_vote, duration_rank, perc_duration_rank, and avg_cumul_duration. 

```{r}
# your code here
usa_early90_high %>%
  mutate(duration_rank = min_rank(duration),
         perc_duration_rank = percent_rank(duration),
         avg_cumul_duration = cummean(duration)) %>%
  filter(perc_duration_rank > 0.5 & perc_duration_rank < 0.6) %>%
  select(title, year, duration, avg_vote, duration_rank:avg_cumul_duration)
```

# The transmute function

The `transmute()` function in dplyr works in a similar way to `mutate()`, but it drops all columns *except* those it has just made. 

```{r}
# use transmutate
messi_career %>%
  transmute(cumul_goals = cumsum(Goals),
         mean_cumul_app = cummean(Appearances))
```


The behaviour of transmute can be helpful in certain situations, but if you really want to keep some columns, you can add them into the transmute statement. For example, in the example below I might want to keep the Goals and Appearances columns for comparison with the cumulativate calculations I've made. 

```{r}
# keep Goals and Appearances
messi_career %>%
  transmute(cumul_goals = cumsum(Goals),
         mean_cumul_app = cummean(Appearances),
         Goals, 
         Appearances)
```
## Transmute exercise

```{r}
movie_durations <- movies_imdb %>%
  transmute(avg_duration = cummean(duration),
            avg_duration_hours = avg_duration*0.0166667,
            duration,
            year) %>%
  filter(avg_duration < median(avg_duration))

movie_durations %>%
  filter(year == round(mean(year)))
```


# Change column names

Changing column names is a very useful part of data science. Sometimes you'll get a dataset with column names that are not very meaningful, or far too long. There are a few methods for changing column names, with the easiest (not surprisingly) being the tidyverse solution. 

The first step in changing column names is actaully viewing what the names are! Two functions in R exist for this: `colnames()` and `names()`. They do the same thing...so I prefer `names()` as it is less typing. 
```{r}
# view a datasets column names
colnames(messi_career)
names(messi_career)
```

The non-tidyverse way of changing column names is to use the `names()` function. If you are changing one column you use indexing using `[]`, and multiple columns you use `c(). 
```{r}
# Make a data frame
df <- data.frame(
  column1 = rep("Hello", 4),
  column2 = sample(1:10, 4),
  column3 = seq(1:4),
  integer = 4:7,
  factor = factor(c("dog", "cat", "cat", "dog"))
)

df

# multiple columns using names
names(df) <- c("string", "random", "sequence", "integer", "factor")
names(df)

# using names and number index
names(df)[1] <- "a_string"
names(df)

# using logic and names
names(df)[names(df) == "sequence"] <- "its_a_sequence"
names(df)
```

The main issue with these techniques is 1) it can get really messy if you need to rename lots of columns in a larger dataset. 2) I have to rename all my columns if renaming more than one column, otherwise it doesn't work! 3) The syntax is a bit messy, especially the last example

The `rename()` function from dplyr allows for simple changing of column names with no fuss, and solves these problems. 

The syntax is the same as the `mutate()` function, where we have the name of the column we want to make, then what column we are changing: `data %>% rename(new_column_name = old_column_name)`. 

```{r message=FALSE}
# load dplyr
library(dplyr)

# Make a data frame
df <- data.frame(
  column1 = rep("Hello", 4),
  column2 = sample(1:10, 4),
  column3 = seq(1:4),
  integer = 4:7,
  factor = factor(c("dog", "cat", "cat", "dog"))
)

df

# rename columns that need renaming
df_new_col <- df %>%
  rename(string = column1,
         random = column2,
         sequence = column3) 

df_new_col
```

## Rename columns exercise

Let's have a practice renaming some columns in the movies_imdb dataset. 

1) Type in and run `names(movies_imdb)` to get the column names of your dataset. This is a nice way to finding the column names, making it easy to copy and paste the names should you need to
2) Using the `rename()` function from dplyr, change `reviews_from_users` to `User_reviews` and `reviews_from_critics` to `Critic_reviews`
3) Save the result back to `movies_imdb`
4) Type in and run `names(movies_imdb)` again to view the new column names

```{r}
# your code here
names(movies_imdb)

movies_imdb <- movies_imdb %>%
  rename(User_reviews = reviews_from_users,
         Critic_reviews = reviews_from_critics)

names(movies_imdb)
```


# Tidy column names with janitor

Sometimes you have a dataset that has messy or ugly column names. 

It changes all the column names to the same format.
