---
title: "Extras"
author: "Andrew Moles"
date: "04/11/2020"
output: html_document
---

# String manipulation with grep and gsub
It is often useful to be able to pattern match or replace values in strings. In R there are two very useful functions for this: `grep()` and `gsub()`. 

## String matching with grep
We use `grep()` for string matching. We give it the string or part of string we are looking for, and it will return the indexes in the vector for the selected string/s.
```{r}
places <- c(rep("Hampshire", 2), rep("London", 5), rep("Kent", 1), rep("Surrey", 3))
grep("London", places)
```
It is very useful for indexing strings.
```{r}
places[grep("London", places)]
```

## grep exercise
We have some patient IDs. We want to just pull out Id's 1, and 10 to 19. So you should get *"ID_1"  "ID_10" "ID_11" "ID_12" "ID_13" "ID_14" "ID_15" "ID_16" "ID_17" "ID_18" "ID_19"*

1)  Using `grep()`, index the ID variable to pull out the requested ID's and assign it to new variable called newID.
2)  Print your newID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String replacement with gsub
With `gsub()` we give the function the pattern we are looking to replace, what to replace it with, and the variable or vector to work on.
```{r}
Names <- c("Andrew", "Andrea", "Angela")
gsub("An" ,"" , Names)
```
Here we are removing the An from the names in the Names vector. 

## gsub exercise
Using the same IDs variable we are going to use gsub to replace the *ID* from each string and replace it with *patientID*. You should end up with IDs that looks like *patientID_3*. 

1)  Using gsub replace ID with patientID, assign the results to a variable called patientID.
2)  Print your new patientID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String manipulation exercise
For this exercise I have given you the code but it is in the wrong order. You need to re-arrange the code so it runs correctly. Comment on what each line of code is doing. 

The end result you are aiming for is: **"These 4 pokemon have 'ar' in their names: Charmander, Charmeleon, Charizard, Wartortle"**
```{r eval=FALSE}
pokemon <- gsub("[0-9]", "", pokemon)

paste0("These ",arPokes_num, " pokemon have 'ar' in their names: ", arPokes)

arPokes_num <- sum(length(arPokes))

arPokes <- pokemon[grepl("ar", pokemon)]

pokemon <- c("Bulbasaur001", "Ivysaur002", "Venusaur003",
             "Charmander004", "Charmeleon005", "Charizard006",
             "Squirtle007", "Wartortle008", "Blastoise009")

arPokes <- paste(arPokes, collapse = ", ")
```


# Tabulation using factors

It can be helpful to know how many occurrences (counts) of each category you have in a factor. You can do this using the `table()` function, which allows us to do simple tabulations. 
```{r}
# colours vector
cols <- c('red','blue','red','green','red','green',
          'blue','pink','green','red','red','blue')
# tabulation of colours vector
table(cols)
```

The `table()` function allows us to do calculations such as percentages. We have used the `length()` function to tell R how much to divide by as opposed to adding the number of variables in cols (12). This is good practice as the data used can change. We can also use the `prop.table()` function as shown below. 
```{r}
# colours vector
cols <- c('red','blue','red','green','red','green',
          'blue','pink','green','red','red','blue')
# found out the percentages rather than counts
table(cols)/length(cols)*100
# found out the percentages using prop.table
prop.table(table(cols))*100
```

***
***

## Tabulation exercise

1)  Make the StringsOrchestra variable, defined below, into a factor. 
2)  Using `table()` work out the percentage each instrument makes up of this orchestra. Test this out with both the examples used above. 
```{r}
StringsOrchestra <- c(rep('first Violin',12),
                      rep('second Violin',10),
                      rep('Viola',8),
                      rep('Cello',6),
                      rep('Double bass',4))

# your code here

```

# Adjusting column names

To just look at the column names for your data frame there are two functions you can use: `names()` and `colnames()`. As `names()` is shorter and easier to type this is the better option. To run `names()` simply give it your data frame as shown below.

```{r}
#names()
```

For the example data we have been using, I want to change string to string1 and integer to integer1. There are two main ways of changing the name of a column. The first is using the columns index number, in this case integer is 2. This method is not recommended, as the size of your data frame can change.

```{r}
# using number index
#names(exampleDat)[2] <- "integer1"
#names(exampleDat)
```

The recommended method is below, here we call names, and within names we index our data. Within the index we call names again with the data and look for column names that are equal to `string`. It will search for the exact match so is case sensitive. 

In session 5 we will be covering conditional statements like these in detail. For now, run the code to test it out: 
```{r}
# recommended using logic
#names(exampleDat)[names(exampleDat) == "string"] <- "string1"
#names(exampleDat)
```
 
## Column names task

We've been recording our running times and distances and have put them in a data frame. However, the column names we have used are not very informative. Using the examples above change the column names as follows:

1)  Change runTime to runTime_mins
2)  Change Len to distance_km
3)  Change speed to minsPerKm
4)  What function would we use to get the mean, max, min etc. on the dataset? Run that now to review your data. 
```{r}
# running data frame
data <- data.frame(runTime = sample(20:40, 10),
                   Len = sample(1:10, 10),
                   week = as.factor(c(rep("weekOne",3), rep("weekTwo", 4), rep("weekThree", 3))))
# work out your minutes per kilometer
data$speed <- data$runTime/data$Len

# your code here


```
*Note: The `sample()` function takes a sample from a distribution. In our case for runTime it picks out 10 variables randomly from numbers between 20 to 40.*

# Introducing the here package

So far we have been loading files by using file paths. Another option is using the here package. 

Short intro to here, example, and task. move here stuff from earlier up. Show both ways. 

Here is good for many reasons, it can build a file path that is not dependent on the operating system you are using. For example, file paths in Mac use forward slashes (/) and Windows use backward slashes (\). This is great for using your code across different systems, and for sharing your code.

```{r}
# example using here to write a file
```

## Here package task

We can direct the `list.files` function to the data file where our data is now saved using the `here` package. If you are not sure if your data file is in the same directory as this notebook run the below command to list the files in your directory. *note: you will likely have to change the file names to the names of your files, e.g. Workshop4b*.  

`list.files(here("datafolder", "datafile"))`

*not working for me, they just need to put in data*
```{r}
library(here)
list.files(here("Workshop4b", "data"))
list.files("data/")
```
This command should print out: *pokemonGen1.csv*, *pokemonGen1.xlsx*, *pokemonGen2.csv*, *pokemongen2.xlsx*.

We can also call packages using the package name and two colons (::), see example below:
```{r}
list.files(here::here("Workshop4b", "data"))
```

# writing multiple sheets to excel

In this coding challenge you will need to re-organise the code to get it to run. The tfl cycles dataset is a bit messy, the first two columns as the raw data, the rest is aggregated data. We are going to separate it out into separate data frames to make it easier to read. 


```{r}
# load libraries
library(readxl)
library(writexl)

# load in tfl cycle data (might need to change file path)
tflCycle <- read_xlsx("data/tfl-daily-cycle-hires.xlsx", sheet = 2) 

# Take out raw data (hire by day)
tflCycleByDay <- tflCycle[,1:2]

# hire by month and ave hire time
tflCycleByMonth <- tflCycle[,c("Month...4","Number of Bicycle Hires...5","Average Hire Time (mins)")]

# hire by year
tflCycleByYear <- tflCycle[1:10,c("Year", "Number of Bicycle Hires...8")]

# most popular day, months and year
maxDay <- tflCycleByDay[which.max(tflCycleByDay$`Number of Bicycle Hires...2`),]
maxMonth <- tflCycleByMonth[which.max(tflCycleByMonth$`Number of Bicycle Hires...5`),]
maxHire <- tflCycleByMonth[which.max(tflCycleByMonth$`Average Hire Time (mins)`),]
maxYear <- tflCycleByYear[which.max(tflCycleByYear$`Number of Bicycle Hires...8`),]

# making month a character for paste
maxHire$Month...4 <- as.character(maxHire$Month...4)

# pasting an output
paste("The highest highest average cycle hire time per month was", maxHire[1,3], 
      "minutes, with the total hires of", maxHire[1,2], "on the", maxHire[1,1],
      "just after lockdown started!")

# Make list of data frames
sheets <- list("by_day" = tflCycleByDay, "by+month" = tflCycleByMonth,
               "by_year" = tflCycleByYear) 
# Write out the data into xlsx file, with separate sheets
write_xlsx(sheets, "data/tflCycleHire.xlsx")
```


# Extra: Sampling explaination

If you are curious about how the `sample()` function works, which we used earlier, run the below code several times, reviewing the histogram it produces. This process is known as sampling with replacement, which means from our sample (numbers 1 to 10) we randomly extract a set number of values (20 in our example) which can be duplicates (extract the same number more than once).

```{r}
# sample with replacement
rand_num <- sample(1:10, 20, replace = TRUE)
# plot using hist function
hist(rand_num, breaks = 0:10)
# add line for mean
abline(v=mean(rand_num),col="darkblue")
```

Sampling without replacement means we don't extract the same number more than once, this means the number we extract must be less than the amount of data in our sample; in our second example we have 1 to 10 and are extract 5 values.

```{r}
# sample without replacement
rand_num2 <- sample(1:10, 5, replace = FALSE)
# plot using hist function
hist(rand_num2, breaks = 0:10)
# add line for mean
abline(v=mean(rand_num2),col="darkgreen")
```


# Useful datasets

School workforce data:
https://explore-education-statistics.service.gov.uk/find-statistics/school-workforce-in-england#releaseHeadlines-charts


# Model outputs are just lists

Save for analysis workshops
```{r}
ronaldo <- read.csv("https://raw.githubusercontent.com/andrewmoles2/webScraping/main/R/data/ronaldo_club.csv")

ronaldo_lm <- lm(total_apps ~ total_goals, data = ronaldo)
sum_ron <- summary(ronaldo_lm)
ronaldo_lm$model
sum_ron$coefficients
sum_ron$adj.r.squared
```

# Accessing and working with data frames from lists

Save for the apply and looping session
```{r}
# vector with several (named) urls
urls <- c(dinosaurs = "https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/r-fundamentals-7/data/dinosaurs.csv",
  pokemon = "https://raw.githubusercontent.com/andrewmoles2/webScraping/main/R/data/pokemon.csv", 
  ronaldo = "https://raw.githubusercontent.com/andrewmoles2/webScraping/main/R/data/ronaldo_club.csv",
  messi = "https://raw.githubusercontent.com/andrewmoles2/webScraping/main/R/data/messi_club.csv")

# lapply to load and put into a list
data_files <- lapply(urls, read.csv)

str(data_files)
```

Note you can do this with any file type, such as a csv files you have saved in a folder of your computer. See this stackoverflow thread for examples: <https://stackoverflow.com/questions/11433432/how-to-import-multiple-csv-files-at-once>

# Why is data visualisation important?

```{r}
library(datasauRus)
library(ggplot2)
library(dplyr)

datasaurus_dozen %>%
  filter(dataset == "dino") %>%
ggplot(aes(x, y)) +
  geom_point()

(avg <- datasaurus_dozen %>%
  group_by(dataset) %>%
  summarise(mean_x = mean(x),
            mean_y = mean(y)))

datasaurus_dozen %>%
  ggplot(aes(x, y)) +
  geom_point() +
  facet_wrap("dataset") +
  geom_hline(yintercept = avg$mean_y, colour = "grey") +
  geom_vline(xintercept = avg$mean_y, colour = "grey")

datasaurus_dozen %>%
  filter(dataset == "dino" | dataset == "star" | 
           dataset == "dots" | dataset == "circle") %>%
  ggplot(aes(x, y)) +
  geom_point() +
  facet_wrap("dataset") + 
  geom_hline(yintercept = avg$mean_y, colour = "grey") +
  geom_vline(xintercept = avg$mean_y, colour = "grey")
```


