---
title: "Extras"
author: "Andrew Moles"
date: "04/11/2020"
output: html_document
---

# String manipulation with grep and gsub
It is often useful to be able to pattern match or replace values in strings. In R there are two very useful functions for this: `grep()` and `gsub()`. 

## String matching with grep
We use `grep()` for string matching. We give it the string or part of string we are looking for, and it will return the indexes in the vector for the selected string/s.
```{r}
places <- c(rep("Hampshire", 2), rep("London", 5), rep("Kent", 1), rep("Surrey", 3))
grep("London", places)
```
It is very useful for indexing strings.
```{r}
places[grep("London", places)]
```

## grep exercise
We have some patient IDs. We want to just pull out Id's 1, and 10 to 19. So you should get *"ID_1"  "ID_10" "ID_11" "ID_12" "ID_13" "ID_14" "ID_15" "ID_16" "ID_17" "ID_18" "ID_19"*

1)  Using `grep()`, index the ID variable to pull out the requested ID's and assign it to new variable called newID.
2)  Print your newID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String replacement with gsub
With `gsub()` we give the function the pattern we are looking to replace, what to replace it with, and the variable or vector to work on.
```{r}
Names <- c("Andrew", "Andrea", "Angela")
gsub("An" ,"" , Names)
```
Here we are removing the An from the names in the Names vector. 

## gsub exercise
Using the same IDs variable we are going to use gsub to replace the *ID* from each string and replace it with *patientID*. You should end up with IDs that looks like *patientID_3*. 

1)  Using gsub replace ID with patientID, assign the results to a variable called patientID.
2)  Print your new patientID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String manipulation exercise
For this exercise I have given you the code but it is in the wrong order. You need to re-arrange the code so it runs correctly. Comment on what each line of code is doing. 

The end result you are aiming for is: **"These 4 pokemon have 'ar' in their names: Charmander, Charmeleon, Charizard, Wartortle"**
```{r eval=FALSE}
pokemon <- gsub("[0-9]", "", pokemon)

paste0("These ",arPokes_num, " pokemon have 'ar' in their names: ", arPokes)

arPokes_num <- sum(length(arPokes))

arPokes <- pokemon[grepl("ar", pokemon)]

pokemon <- c("Bulbasaur001", "Ivysaur002", "Venusaur003",
             "Charmander004", "Charmeleon005", "Charizard006",
             "Squirtle007", "Wartortle008", "Blastoise009")

arPokes <- paste(arPokes, collapse = ", ")
```


# Tabulation using factors

It can be helpful to know how many occurrences (counts) of each category you have in a factor. You can do this using the `table()` function, which allows us to do simple tabulations. 
```{r}
# colours vector
cols <- c('red','blue','red','green','red','green',
          'blue','pink','green','red','red','blue')
# tabulation of colours vector
table(cols)
```

The `table()` function allows us to do calculations such as percentages. We have used the `length()` function to tell R how much to divide by as opposed to adding the number of variables in cols (12). This is good practice as the data used can change. We can also use the `prop.table()` function as shown below. 
```{r}
# colours vector
cols <- c('red','blue','red','green','red','green',
          'blue','pink','green','red','red','blue')
# found out the percentages rather than counts
table(cols)/length(cols)*100
# found out the percentages using prop.table
prop.table(table(cols))*100
```

***
***

## Tabulation exercise

1)  Make the StringsOrchestra variable, defined below, into a factor. 
2)  Using `table()` work out the percentage each instrument makes up of this orchestra. Test this out with both the examples used above. 
```{r}
StringsOrchestra <- c(rep('first Violin',12),
                      rep('second Violin',10),
                      rep('Viola',8),
                      rep('Cello',6),
                      rep('Double bass',4))

# your code here

```

# Adjusting column names

To just look at the column names for your data frame there are two functions you can use: `names()` and `colnames()`. As `names()` is shorter and easier to type this is the better option. To run `names()` simply give it your data frame as shown below.

```{r}
#names()
```

For the example data we have been using, I want to change string to string1 and integer to integer1. There are two main ways of changing the name of a column. The first is using the columns index number, in this case integer is 2. This method is not recommended, as the size of your data frame can change.

```{r}
# using number index
#names(exampleDat)[2] <- "integer1"
#names(exampleDat)
```

The recommended method is below, here we call names, and within names we index our data. Within the index we call names again with the data and look for column names that are equal to `string`. It will search for the exact match so is case sensitive. 

In session 5 we will be covering conditional statements like these in detail. For now, run the code to test it out: 
```{r}
# recommended using logic
#names(exampleDat)[names(exampleDat) == "string"] <- "string1"
#names(exampleDat)
```
 
## Column names task

We've been recording our running times and distances and have put them in a data frame. However, the column names we have used are not very informative. Using the examples above change the column names as follows:

1)  Change runTime to runTime_mins
2)  Change Len to distance_km
3)  Change speed to minsPerKm
4)  What function would we use to get the mean, max, min etc. on the dataset? Run that now to review your data. 
```{r}
# running data frame
data <- data.frame(runTime = sample(20:40, 10),
                   Len = sample(1:10, 10),
                   week = as.factor(c(rep("weekOne",3), rep("weekTwo", 4), rep("weekThree", 3))))
# work out your minutes per kilometer
data$speed <- data$runTime/data$Len

# your code here


```
*Note: The `sample()` function takes a sample from a distribution. In our case for runTime it picks out 10 variables randomly from numbers between 20 to 40.*
