---
title: "Extras"
author: "Andrew Moles"
date: "04/11/2020"
output: html_document
---

# String manipulation with grep and gsub
It is often useful to be able to pattern match or replace values in strings. In R there are two very useful functions for this: `grep()` and `gsub()`. 

## String matching with grep
We use `grep()` for string matching. We give it the string or part of string we are looking for, and it will return the indexes in the vector for the selected string/s.
```{r}
places <- c(rep("Hampshire", 2), rep("London", 5), rep("Kent", 1), rep("Surrey", 3))
grep("London", places)
```
It is very useful for indexing strings.
```{r}
places[grep("London", places)]
```

## grep exercise
We have some patient IDs. We want to just pull out Id's 1, and 10 to 19. So you should get *"ID_1"  "ID_10" "ID_11" "ID_12" "ID_13" "ID_14" "ID_15" "ID_16" "ID_17" "ID_18" "ID_19"*

1)  Using `grep()`, index the ID variable to pull out the requested ID's and assign it to new variable called newID.
2)  Print your newID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String replacement with gsub
With `gsub()` we give the function the pattern we are looking to replace, what to replace it with, and the variable or vector to work on.
```{r}
Names <- c("Andrew", "Andrea", "Angela")
gsub("An" ,"" , Names)
```
Here we are removing the An from the names in the Names vector. 

## gsub exercise
Using the same IDs variable we are going to use gsub to replace the *ID* from each string and replace it with *patientID*. You should end up with IDs that looks like *patientID_3*. 

1)  Using gsub replace ID with patientID, assign the results to a variable called patientID.
2)  Print your new patientID variable.
3)  Comment your code.
```{r}
ID <- paste0("ID_", seq(1:50))

# your code here

```

## String manipulation exercise
For this exercise I have given you the code but it is in the wrong order. You need to re-arrange the code so it runs correctly. Comment on what each line of code is doing. 

The end result you are aiming for is: **"These 4 pokemon have 'ar' in their names: Charmander, Charmeleon, Charizard, Wartortle"**
```{r eval=FALSE}
pokemon <- gsub("[0-9]", "", pokemon)

paste0("These ",arPokes_num, " pokemon have 'ar' in their names: ", arPokes)

arPokes_num <- sum(length(arPokes))

arPokes <- pokemon[grepl("ar", pokemon)]

pokemon <- c("Bulbasaur001", "Ivysaur002", "Venusaur003",
             "Charmander004", "Charmeleon005", "Charizard006",
             "Squirtle007", "Wartortle008", "Blastoise009")

arPokes <- paste(arPokes, collapse = ", ")
```
