---
title: "R Intermediate 3: Building functions"
author:
   - name: "Andrew Moles & Francisco Alessandri"
     affiliation: Digital Skills Lab
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    theme: readable
    highlight: pygments
    keep_md: no
    code_download: true
    toc: true
    toc_float: 
      collapsed: false
---

# Objective of workshop

To start making your own functions to use for your data cleaning and analysis in R.

# What this workshop will cover

In this workshop, the aim is to cover how to start making your own functions. We will be covering:

-   Introduce the basics of making functions
-   How to use functions you've made
-   Some general use cases to get you started
-   Using functions with Data Frames: The apply family

------------------------------------------------------------------------

# Functions 101

Through your learning of R you will likely have used hundreds of functions. Building your own functions will allow you to start to make your code more streamlined, as well as apply some automation.

Functions are great at reducing your workload in the long run. Imagine you are an analyst and have to repeat some analysis steps over and over again. You could copy and paste your code several times, which is error prone. Or you could wrap up that code you are repeating into a function, saving you from copying and pasting lots of code!

Functions have a basic structure: `name_of_function <- function(argument) {some code}`

Within the `()` we have arguments. Function arguments are the real values passed to (and received by) the function. Function parameters are the names listed in the function definition.

Basic example, have a function that prints out a welcome message.

```{r}
hello <- function(name) {
  # our code that uses name argument
  message <- paste0("Hello there ", name, ", welcome to R!")
  # return ensures function gives output
  return(message)
}

hello("Andrew")
```

The `return()` function is important, as it makes sure an output is given when the function is used. If we remove it nothing will happen. Run this code chunk to test it out.

```{r}
hello2 <- function(name) {
  # our code that uses name argument
  message <- paste0("Hello there ", name, ", welcome to R!")
}

hello2("Andrew")
```

You can also add in default arguments to your functions. This is useful if you know what the argument will be most of the time. To add default arguments we would assign a value to our argument, in our example this would be `name = "Andrew"`.

```{r}
hello <- function(name = "Andrew") {
  # our code that uses name argument
  message <- paste0("Hello there ", name, ", welcome to R!")
  # return ensures function gives output
  return(message)
}


hello()
```

## Functions 101 exercise

We are going to build a message that welcomes you according to what time of day it is, how pleasant!

Re-write the code below to be a function, using the examples above to help you out. Make sure to get the function to use your name when welcoming you.

```{r, error=TRUE}
name <- "Andrew"
hour <- format(Sys.time(), format = "%H")
hour <- as.numeric(hour)
morning <- 0:11
afternoon <- 12:16
evening <- c(17:23)
day_hrs <- rep(x = c("Morning", "Afternoon", "Evening"), 
               times = c(length(morning), length(afternoon), length(evening)))
names(day_hrs) <- as.character(0:23)
welcome <- paste0("Good ", day_hrs[hour], ", ", name, "!")

welcome

# your code here


```

# Adding more arguments

A function can have several arguments. This is particularly helpful as your functions become more complex.

An example can be calculating the Body Mass Index (BMI) of a person, considering their weight in Kilograms and height in meters, with the formula: $$BMI = \frac{weight}{Height^2}$$

```{r}
BMI <- function(weight, height) {
  weight <- weight
  height <- height
  
  BMI <- weight/(height^2)
  
return(BMI)
  
}

BMI(85, 1.85)
```

## Adding more arguments exercise

Take the code below, and write it into a function. The code finds the percentage of occurrences of a given phrase, regardless of case. We will use a example dataset about dinosaurs. In this example we find the percentage of occurrences of the phrase *saurus* in the dinosaur names column.

Your function should have three arguments: data, column name, and the phrase. Try and add default parameters for col and phrase.

Test out your function on the dinosaur data with the following columns and phrases:

-   column = "Name", phrase = "saurus"
-   column = "Name", phrase = "titan"
-   column = "Class", phrase = "poda"

```{r}
dinosaurs <- read.csv("https://raw.githubusercontent.com/andrewmoles2/rTrainIntroduction/master/r-fundamentals-7/data/dinosaurs.csv")

(sum(grepl("saurus", dinosaurs$Name, ignore.case = TRUE)) / nrow(dinosaurs)) * 100

# your code here



```

# Short hand functions

Functions can be written in short hand, which technically means you do not use the curly brackets `{}` when defining the function. This works well if you have very little code, but is not recommended if you have more than one line of code.

Let's look at some examples. First, let's take our hello function from earlier and make it short hand.

```{r}
hello3 <- function(name) paste0("Hello there ", name, ", welcome to R!")
hello3("Andrew")
```

This works well for this function as it is only one line of code. If the function got more complex we should put all the code into curly brackets `{}`.

# Short hand functions exercise

For this exercise we will be convert the function below into to a short hand function. This simple function counts the number of NA values in your data.

```{r}
# test data
x <- sample(c(1:12, NA, NA))
# function to count missing values
count_missing <- function(df) {
  sum(is.na(df))
}
# test
count_missing(x)
```

Now use the code chunk below to convert the function to a short hand function.

```{r}
# your code here


```

# loops and conditional statements in functions

An advantage of the use of functions is that they can be combined with loops and conditional statements (if and else), simplifying operations that, if repeated, could make our code to messy and hard to understand.

Our function will start from the beginning every time we call it, iterating through the conditional statements. For example, we can use a function to find the factorial of a number.

```{r}
factorial <- function(num) {
  fact <- 1
  
  if(num < 0) {
      return("factorial is only defined for non-negative integers")
  } else if(num == 0) {
      return(fact)
  } else {
      for(i in 1:num) {
        fact <- fact * i
    }
  }
  return(fact)
}

factorial(3)
factorial(6)
factorial(-1)
```

## Loops and conditional statements exercise

In this exercise, we will be identifying Prime numbers. A Prime number is a number such that it is only divisible by itself and one.

Here we present a function that should return a statement to tell you if your number is a Prime number or not. It uses conditional statements and loops.

Your job is the fix the code to get it running.

```{r, error=TRUE}
its_prime <- function(num) 
  count = 0
  if(nu > 2) {
    for( i in 2:(num - 1)) {
      if(num %% i == 0) count = cout + 1
    }
  }
  if(num == 1) {
    count = count + 1
  }
  if(count = 0) {
    return(print(paste0(num, " is a prime number!")))
  }
  return(return(print(paste0(num, " is not a prime number!"))))
}

its_prime(2)
its_prime(15)
```

## BMI exercise

We can now connect both multiple arguments and conditional statements for the next task. You will notice that BMI calculation can be done with different metrics.

Knowing that:

-   1 meter (m) is 100 centimetres (cm)
-   1 meter (m) is 3.28 feet (ft)

Rewrite the BMI function with an extra argument called unit (with a default parameter of meters). Your function will be able to calculate the BMI with any of meters, cm, or feet units of height, and indicate there is a problem if the unit chosen is not valid.

*Hint: A useful strategy to build your own functions is to first write your code as usual, make sure it works, and after that transform it into a function*

```{r}
# your code here


```

# Issues while working with vectors of elements and the Apply family

One of the many potential uses of functions is to automate an operation for a vector of elements, or columns of a Data Frame. This is useful as often you need to do the same operation for every row in your data frame, and it can be tedious, messy, and time-consuming.

Many built-in functions, like the sum or the multiplication operators, do operations *row-wise* in your data frame, and they can even be saved in the same data frame, like this example

```{r}
# make some data
a <- c(2, 5, 11, 4, 7)
b <- c(23, 41, 32, 58, 26)
df <- data.frame(a, b)

df

# calculate sum and product row-wise
df$sum<- df$a + df$b
df$prod <- df$a * df$b

df
```

As you can see, to add a new column we only needed to define it and do the operation. However, if we try to use a function defined by us, like `digit_number()`, a function that returns the number of digits of any number, we may face some difficulties.

```{r, error=TRUE}

digit_number <-function(num){
  if(num < 10) {
    return("1 digit")
  } else if(num < 100) {
    return("2 digits")
  } else if(num < 1000) {
    return("3 digits")
  } else {
    return("4 or more digits")
  }
}

name = c("Andrew", "Michael", "Alexia", "Sue", "Mark")
fav_number = c(19, 9, 37, 21, 4)
people <- data.frame(name, fav_number)

people$n_digit <- digit_number(people$fav_number)

```

This function give us an error of *the condition has length \> 1*. An `if()` statement can only check one element in a vector at one time, and cannot check the whole vector.

To solve this issue, we can use a for loop which can apply the `if()` to every element in the vector.

In the example below we create an empty vector, a loop that goes through every row of your data frame, applying the function in the column `fav_number` and saving it in your new vector. Finally, we add the vector you created to your data frame as the column "n_digit".

```{r}
n_digit <- c()
for(i in 1:nrow(people)){
  value <- digit_number(people[i,]$fav_number)
  n_digit <- c(n_digit, value)
}

people$n_digit <- n_digit
people
```

However, this solution is complicated, requires a lot of code, and can take a long time if performed in larger data frames. To have better implementations of functions in Data Frames, we use the apply family that we learnt about in previous workshops.

In this case, we will use `sapply()`, a function that can be used on elements like data frames, lists, and vectors, and return a vector, or the most simplified output possible. It takes the form `sapply(X, FUN)`, where X is your list, vector or data frame, and FUN is your function.

For example, to solve our digits problem you would do the following:

```{r}
people$n_digit <- sapply(people$fav_number, digit_number)
people
```

Simpler, faster and tidier than a loop! You can remember an important characteristic of the apply family: you should put only the name of your function `digit_number`, not calling it with the parenthesis `digit_number()`, take care of this, as it is one of the main reasons why your code could not work.

## Exercise apply, Data frames, and functions

The following data frame, called family, contains the information of an extended family group, such as the name, gender, and the birth year.

The objective is to create a new column called "generation" that contains the generation (Gen Z, Millennial, etc.) of each family member. To calculate that, consider the following classification of generations:

-   Gen Z: Born in 1997 or after
-   Millennial: Born between 1981 and 1996
-   Gen Z: Born between 1965 and 1980
-   Boomer II: Born between 1955 and 1964
-   Boomer I: Born between 1946 and 1954
-   Post war: Born between 1928 and 1945
-   WW II: Born between 1922 and 1927
-   If the information is wrongly coded or any year not in this categories, the function should return "Unknown"

To do this you will first have to make a function, then use an apply function, such as `sapply()`, to *apply* your function to each row in `family`, assigning the result as a new column in your `family` data frame. 

*Hint:* our year vector has NA values. To ensure our if statement works we need to include `is.na()` with a statement like `birth_year >= 1900 & !is.na(birth_year)`. This checks for and ignores missing values in the statement.

```{r}
# family data frame
name <- c("Maria", "John", "Elisa", "Charles", "Margaret", "Andrew", "Rosa", "Zain", "Harry", "Yujong", "Peter", "Marcus", "Edward", "George", "Caroline", "Ruth")
gender <- c("F", "M", "F", "M", "F", "M", "F", "M", "M", "M", "M", "M", "M", "M", "F", "F")
year <- c(1988,2004,1996,1946,1966,1941,NA,1940,1953,1985,1972,NA,2001,1945,1957,1925)

family <- data.frame(name, gender, year)

# Your code here


```

As you can see functions can be really useful to make and use for your own use cases! In the next workshop, we will go into more depth with functions, covering more complex functions we can build.

# Useful resources

<https://adv-r.hadley.nz/functions.html>

<https://www.w3schools.com/r/r_functions.asp>

# Final task - Please give us your individual feedback!

We would be grateful if you could take a minute before the end of the workshop so we can get your feedback!

<https://lse.eu.qualtrics.com/jfe/form/SV_6eSrOVWuit28qcS?coursename=R%Intermediate%3:%Making%functions&topic=R&prog=DS&version=23-24&link=https://lsecloud.sharepoint.com/:f:/s/TEAM_APD-DSL-Digital-Skills-Trainers/EspgYj7sxEJOlVQjxrPbE8kBRxeicoivL5aRpzF7zMet8g?e=7FkAHg>

The solutions we be available from a link at the end of the survey.

# Individual coding challenge

Lets do a more realistic exercise!

You have a friend who is a GP. She has a list of people that are assigned to her, and she has data for them including their last measured weight, height, and the unit of height measured.

She has received the news that people with obesity have to urgently to get a new COVID-19 vaccine booster. She asks for your help to automatise that task, as she works with thousands of patients and doing manually is too tiring.

She provided some example data to help you write some code for her. Run the code chunk below to make and review the data.

```{r}
# build gp data frame
Names <- c("Noah", "Oliver", "George", "Arthur", "Muhammad", "Leo", "Harry", "Oscar", "Olivia", "Amelia", "Isla", "Ava", "Ivy", "Freya", "Lily", "Florence")
Gender <- c("M", "M", "M", "M", "M", "M", "M", "M", "F", "F", "F", "F", "F", "F", "F", "F")
Weight <- c(96, 81, 87, 83, 95, 98, 76, 83, 59, 51, 71, 69, 66, 90, 65, 81)
Height <- c(177, 1.75, 1.88, 5.77, 188, 1.80, 1.71, 1.81, 5.28, 1.63, 160, 1.72, 1.74, 1.57, 148, 1.56)
Unit <- c("cm", "m", "m", "ft", "cm", "m", "m", "m", "ft", "m", "cm", "m", "m", "m", "cm", "m")

GP <- data.frame(Names, Gender, Weight, Height, Unit)
head(GP)
```

1.  Look up in the World Health Organization (WHO) which are the values defined for obesity using people's BMI

2.  Create a function that will return if a person is obese or not (you can use the old BMI calculator you already created)

3.  Using an apply function (or a map function) make a new column called Obesity or similar where you use your new function to calculate if someone in the data is obese or not

4.  Create another function that will detect if a person is obese and print a message that says "Dear PERSON, our information tell us you are eligible for a COVID-19 booster shot, please feel welcome to go to your GP as soon as possible to receive it", with PERSON being the persons name

5.  Using an apply function (or a map function) print out the messages

*Hint:* When you want to use a function with multiple arguments in a data frame, you should look the `mapply()` function. Try `?mapply()` to look up the documentation. If you'd prefer to use the `purrr` library, you'll need either `pmap()` or `map2()`.

```{r}
# your code here


```
