---
title: "R Workshop 1 - Welcome to R!"
author: "Your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

# Using R as a calculator

We can use R to do simple or advanced calculations for us. Remember to run the code press the green play button or press Ctrl + Enter (or Cmd + Enter on Mac).

```{r}
7 * 6
5 / (2^2)
(16 - 4) + (1 * 9)
```

## Exercise (R as calculator) 

Use R to work out the following arithmetic:

1)  44 * 26
2)  9.6/1.6
3)  (12+4)^2
4)  (4*9)/(5+6)
5)  (22/36) * 100

```{r}
# type your code here
```

# Assigning variables

A variable is a named storage of information. In our case today we are storing numbers. 

We can assign variables by using <-. You should see the variable appear to your right in the global enviroment once you've run this command. 
```{r}
height <- 155
```
We can then print the output of the variable by typing in its name. 
```{r}
height
```

## Task

When calling a variable, be careful to type it exactly (you can also copy it or use code completion to help). 

Try calling the height variable, but spell it incorrectly. You should get an error with something like *Error: object 'hieght' not found*

```{r}
# type incorrect height call here
```

## Example

We can do calculations on these variables, just as we did before. We first assign the variables, then use them in the calculation. 

```{r}
# test scores
Score1 <- 42
Score2 <- 92
Score3 <- 68
# average score calculation
AveScore <- (Score1+Score2+Score3)/3
# print average score
AveScore
```

You will have noticed the hashtags (#) with text in the above example. These are called comments. In later R sessions we will use a lot of comments to tell us (and others) what each line or section of code is doing.

```{r}
# this is a comment
```

# Reassigning variables

You can also change the value of a variable you have already assigned. Here we are going to add our new pay check to our previous bank balance. Be sure to run the code.

```{r}
# create variables
BankBalance <- 100
PayCheck <- 250
# add old bank balance and pay check, assigning result to bank balance
BankBalance <- BankBalance + PayCheck
# print bank balance
BankBalance
```

## Task

Try and add another pay check of £50 to the bank balance variable.

1)  Make a variable called PayCheck2 with a value of 50
2)  Assign BankBalance, and calculate BankBalance + PayCheck2
3)  Print the outcome

```{r}
# Reassigning variables task

```

