---
title: "R Workshop 1 - Welcome to R!"
author: "Your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

# Using R as a calculator

We can use R to do simple or advanced calculations for us. Remember to run the code press the green play button or press Ctrl + Enter (or Cmd + Enter on Mac).

```{r}
7 * 6
5 / (2^2)
(16 - 4) + (1 * 9)
```

## Exercise (R as calculator) 

Use R to work out the following arithmetic:

1)  44 * 26
2)  9.6/1.6
3)  (12+4)^2
4)  (4*9)/(5+6)
5)  (22/36) * 100
6)  Remainder of 55/2

```{r}
# type your code here
```

# Assigning variables

A variable is a named storage of information. In our case today we are storing numbers. 

We can assign variables by using <-. You should see the variable appear to your right in the global enviroment once you've run this command. 
```{r}
height <- 155
```
We can then print the output of the variable by typing in its name. 
```{r}
height
```

## Task

When calling a variable, be careful to type it exactly (you can also copy it or use code completion to help). 

Try calling the height variable, but spell it incorrectly. You should get an error with something like *Error: object 'hieght' not found*

```{r}
# type incorrect height call here
```

## Example

We can do calculations on these variables, just as we did before. We first assign the variables, then use them in the calculation. 

```{r}
# test scores
Score1 <- 42
Score2 <- 92
Score3 <- 68
# average score calculation
AveScore <- (Score1+Score2+Score3)/3
# print average score
AveScore
```

You will have noticed the hashtags (#) with text in the above example. These are called comments. In later R sessions we will use a lot of comments to tell us (and others) what each line or section of code is doing.

```{r}
# this is a comment
```

# Reassigning variables

You can also change the value of a variable you have already assigned. Here we are going to add our new pay check to our previous bank balance. Be sure to run the code.

```{r}
# create variables
BankBalance <- 100
PayCheck <- 250
# add old bank balance and pay check, assigning result to bank balance
BankBalance <- BankBalance + PayCheck
# print bank balance
BankBalance
```

## Task

Try and add another pay check of £50 to the bank balance variable.

1)  Make a variable called PayCheck2 with a value of 50
2)  Assign BankBalance, and calculate BankBalance + PayCheck2
3)  Print the outcome

```{r}
# Reassigning variables task

```


## Exercise 1 (Assigning variables)

Use R to work out a body mass index (BMI) of someone who is 79kg, and 1.77m tall.

1)  Assign the variables of weight and height
2)  Assign the variable of BMI, and calculate the BMI based of the weight and height variables
3)  Print the outcome
4)  Add comments on what each line of code is doing

```{r}
# Exercise: BMI calculation

```

## Exercise 2 (Assigning variables)

Debug the code below that is finding the weighted average of a students coursework and exam scores. You should find three errors:

*  logic (maths) error
*  syntax error
*  naming error

```{r eval=FALSE}
# Exercise: weighted average debugging
exam1 <- 52
coursework1 <- 82
exam2 <- 78
coursework2 < 48

cw_weight <- 0.4
ex_weight <- 0.6

course1 <- (exam1 * ex_weight) + (coursework * cw_weight)
course2 <- (exam2 * ex_weight) + (coursework2 * cw_weight)

overall_grade <- (course1 + course2)/3

overall_grade
```

# Exercise 3 (Salary calculation)

Robin and Charlie are a married couple, one gets paid an hourly rate, the other has an annual salary. They want to workout how much annual salary they have combined before tax. Out of interest Charlie also wants to know what her hourly rate is before tax. 

They used simple calculations using the following formulas:  

(number hours worked per week x hourly rate) x number of weeks worked = annual salary
(annual salary ÷ number of weeks in a year) ÷ hours worked per week = hourly rate

Re-arrange the code so the calculations run. You should have both the combined salary and Charlies hourly rate calculations printed. 

```{r eval=FALSE}
# weeks in year
weeksYear <- 52

# Charlies hourly rate
Charlie_HourlyRate <- (Charlie_annualSary / weeksYear)/Charlie_hoursPerWeek
Charlie_HourlyRate

# salaries
Robin_HourlyRate <- 16.5
Charlie_annualSary <- 31800

# Combined salaries
CombinedSalaries <- Robin_annualSalary + Charlie_annualSary
CombinedSalaries

# Robins annual salary
Robin_annualSalary <- (Robin_hoursPerWeek * Robin_HourlyRate) * Robin_weeksWorking

# hours worked
Robin_hoursPerWeek <- 25
Robin_weeksWorking <- 48
Charlie_hoursPerWeek <- 35
```

# Workshop survey

This is the first time that we are exploring a remote learning format for our workshops and we would be grateful if you could take 2 mins before the end of the workshop to get your feedback!

https://lse.eu.qualtrics.com/jfe/form/SV_9zagWkOtzNhmqt7?course=D025-R1NV&topic=R&cohort=MT20

# Individual (take home) coding challenge

A take home coding task for you.

Task: Splitting a Pizza Pilgrims restaurant bill between 3 friends; Roger, Amal and Genevieve.

*  Roger has a Calzone Ripieno (£11), and a San Pellegrino (£2).
*  Amal has a Bufala (£9), and Birra Moretti (£4.50).
*  Genevieve has a Portobello Mushroom & Truffle (£10), and water to drink.
*  Genevieve and Amal also share a Nutella Pizza ring for dessert (£5.5). 


1)  Make a variable for each friend that is the sum of their order (e.g. Roger would be 11 + 2)
2)  Make a variable for the shared food
3)  Make a variable called TotalBill, and calculate their total bill
4)  For those that shared food, add the shared food to their bill
5)  Comment your code
6)  Print the total bill and what each friend owes

```{r}
# individual coding challenge
```

